Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-13 surfaces-write-up/surface-write-up.tex
added saucy intro

\section{Introduction}

@@

\begin{lem}
\label{lem:composite-map}
Let $D = \sum_{i=1}^{n}\alpha_i D_i$ where $D_i = V(f_i)$. Suppose we have a surjection $\phi: \bk[x_1,x_2,\ldots, x_r] \rightarrow R_D, x_i \mapsto p_i(f_1, \ldots, f_n),$ where $p_i$ is a monomial in $f_1,\ldots, f_n$. Then, define
\begin{align*}
	\Sigma = \langle u^d y_1^{c_1} \cdots y_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{} \bidb_i c_i. \rangle 
\end{align*}
In this case, we have a factorization of $\phi$ given by maps $\chi, \psi$ defined by
\todo{Make this have mapsto on bottom}
$$\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}y_1^{c_{i1}} \cdots y_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}
\end{tikzcd}$$
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$ is at most the maximum of the minimal degree of generation of $\ker \chi$ and the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
Note that we have an exact sequence
$$\begin{tikzcd}
0 \ar r & \ker \chi \ar r & \ker \phi \ar r & \ker \psi \ar r & 0
\end{tikzcd}.$$
To see this, note that the left map is an inclusion because $\phi = \chi \circ \psi$. Next, the right map, given by $\chi|_{\ker \phi}:\ker \phi \rightarrow \ker \psi$ is a surjection, again because $\phi = \chi \circ \psi$. The composition of the two maps is $0$ by definition of the kernel of $\chi$. Finally, if $a \in \ker \phi$ and $a$ lies in the kernel of the right map, then by definition $\chi(a) = 0$.
\end{proof}

@@

\todo{possibly change this to only have two equations}
\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ so that 
\begin{align}
\label{align:pm-span}
\sum_{i=0}^{n} c_i \cdot \bida_i \cdot f_i = 0 & \text{ and } &c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}

\noindent
\todo{the above ``and'' needs to be placed better}
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=1}^n f_i^{c_i} $ such that
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor & \text{ for } & i > m
\end{align}

@@

\begin{proof}
By definition of $H^0(\bp^m,dD)$, functions satisfying conditions 
~\eqref{align:pm-span} lie in $H^0(\bp^m,dD)$. To complete the proof, it suffices to check functions satisfying conditions ~\eqref{align:pm-basis} form a basis of $H^0(\bp^m,dD)$. Note that there $\binom{m+ \lfloor dD \rfloor }{m}$ functions satisfying condition ~\eqref{align:pm-basis}. However we know $h^0(\bp^m,dD) = \binom{m+ \lfloor dD \rfloor }{m},$ so it suffices to show that those functions satisfying condition ~\eqref{align:pm-basis} are independent. This follows from the assumption that $f_0,\ldots, f_n$ form a basis of degree 1 rational functions, and so degree $\lfloor d \deg D \rfloor $ monomials in $f_0,\ldots, f_n$ form a basis of degree $\lfloor d \deg D \rfloor $ rational functions.
\end{proof}
\todo{the following bound is the generation bound beginning from evan's paper}

@@

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
$E$ satisfies the conditions of proposition \todo{cite earlier 
proposition}, so by applying that proposition $R_D$ is generated in 
degree at most
\[
	\sigma = \sum_{i\in T_=(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{i\in T_+(D) \atop
	j \in T_-(D)} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}

@@

\begin{proof}
We first claim that $\psi$ has a relation of the form
\begin{equation}\label{eqn:relations-psi}
	(z_i - \beta_j)\prod_{j=1}^n {z_j}^{c_{i}}
\end{equation}

@@

Note that we can view the relations from Equation
\ref{eqn:relations-psi} as an element of $S$ given
by $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$.  Let $S'\subseteq (\mathbb{R}_{\ge 0})^{n+1}$ be the
set of $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$ satisfying $c_i \ge -\alpha_i d$,
$\bida_j + \sum_{i = 1}^m \bida_i c_i = 0$, and $\bidb_j + \sum_{i=1}^n \bidb_i c_i
= 0$.  A careful computation deduces that $\gamma (d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots, c_n)$ is of the form 
\[
	E + \sum_{k\in T_=} r_k e_k + \sum_{h\in T_+ \atop k\in T_-} r_{h,k} e_{h,k},
\] 
where $E$ is $\mathbb{R}_{\ge 0}$
combination of elements in 
\[
	\{e_i \mbox{ if } i\in S\} \cup \{e_{i,j}: j\in S_- \mbox{ if } i \in S_+\} \cup \{e_{j,i}: j\in S_+ \mbox{ if } i \in S_-\}.
\]

Let $M$ be the set of monomials of $\beta_j$.  Then for $\mu\in M$, the corresponding lattice point is of the form
\[
	E_\mu + \sum_{k\in T_=} r_k e_k + \sum_{h\in T_+ \atop k\in T_-} r_{h,k} e_{h,k}
\]
where $E_\mu$ is a $\mathbb{R}_{\ge 0}$-combination of elements in
\begin{align*}
			& \{e_j : j\in T_= \cap \{1, 2, 3, 4\}\} \\
	\cup \; 	& \{e_{j, k}: j\in T_= \cap \{1, 2, 3, 4\}, k\in T_-\}\\
	\cup \; 	& \{e_{k,j}: k\in T_+, j\in T_- \cap \{1, 2, 3, 4\}\}.
\end{align*}

Let $F\in R_D$ correspond to $E + \sum_{k\in T_=} (r_k) e_k + \sum_{h\in T_+ \atop k\in T_-} (r_{h,k}) e_{h,k}$ where $(r)$ signifies the fractional part of $r$.  Similarly, let $F_\mu \in R_D$ correspond to $E_\mu + \sum_{h\in T_+ \atop k\in T_-} (r_{h,k}) e_{h,k}$.  Then $F - \sum_{\mu\in M} F_\mu = 0$ gives us a relation in degree at most
\[
	\sigma + \deg(F),
\]
which generates $(z_i - \beta_i) \prod_{i=1}^n y_i^{c_i}$
We can bound $\deg(F)$ by
\[
	\deg(F) \le \left\{ \begin{split}  \ell_i \gcd(\bida_i, \bidb_i) \mbox{ if } i \in S_= 
			\\ \max_{j\in S_-}(\ell_{i,j} (\bida_i \bidb_
j - \bida_j \bidb_i)) \mbox{ if } i\in S_+ \\
			\\ \max_{j\in S_+}(\ell_{j, i} (\bida_j \bidb_
i - \bida_i \bidb_j)) \mbox{ if } i \in S_- \end{split}  \right.
\]
Therefore, $\ker(\psi)$ is generated in degree at most
\[
	\tau = \sigma
	+ \max( \max_{i\in S_=}(\ell_i \gcd(a_i, b_i)), \max_{i\in S_+ \atop j\in S_-}(\ell_{i,j} (\bida_i \bidb_j - \bida_j \bidb_i))).
\]

>>>>

\section{Introduction}
\todo{Clean up this stuff on siegel modular forms, hilbert modular forms, say something about hasse keel program, canoncical rings studied before}

\todo{Might be reasonable to think about higher dimensional rational varieties that are similar to these hirzebruch surfaces.}

\todo{The following stuff is almost nonsensical now, will be edited much more soon}

Hasse keel program, birational map from $proj R(K + \Delta)$ to $M_2$, they showed finite generation, we're giving explicit generators and bounds.

The work in this paper also yields bounds for certain rings of Hilbert modular forms. Recall that a Hilbert modular form is a generalization of modular forms to multiple variables.

Siegel modular forms are a generalization of modular forms on the Siegel upper half plane to automorphic forms on the Siegel upper half space. Explicitly, the Siegel uper half space is $\mathscr H \cong \bc^n/\Lambda$ and siegel modular forms are functions on $Sp_{2g} \backslash \mathscr H.$ Then, modular functions are differentials invariant under the action of $Sp_{2g}$. In the case that $\Gamma \backslash \mathscr H$.

Example: the Siegel 2-upper half plane $\bc^2/\Lambda$ Note that $\bc^2/\Lambda$ is algebraic if and only if there exists a hermitial form on $\Lambda$. Then, $Sp_{2g}$ acts on algebraic lattices. The quotient by this gives you the space of abelian varieties. This might be an open subset of projective space, according to David, 65$\%$.

@@

\begin{lem}
\label{lem:composite-map}
Let $D = \sum_{i=1}^{n}\alpha_i D_i$ where $D_i = V(f_i)$. Suppose we have a surjection $\phi: \bk[x_1,x_2,\ldots, x_r] \rightarrow R_D, x_i \mapsto p_i(f_1, \ldots, f_n),$ where $p_i$ is a monomial in $f_1,\ldots, f_n$. Then, define
\begin{align*}
	\Sigma = \langle u^d y_1^{c_1} \cdots y_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{} \bidb_i c_i. \rangle 
\end{align*}
In this case, we have a factorization of $\phi$ given by maps $\chi, \psi$ defined by
\todo{Make this have mapsto on bottom}
$$\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}y_1^{c_{i1}} \cdots y_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}
\end{tikzcd}$$
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$ is at most the maximum of the minimal degree of generation of $\ker \chi$ and the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
Note that we have an exact sequence
$$\begin{tikzcd}
0 \ar r & \ker \chi \ar r & \ker \phi \ar r & \ker \psi \ar r & 0.
\end{tikzcd}$$
To see this, note that the left map is an inclusion because $\phi = \chi \circ \psi$. Next, the right map, given by $\chi|_{\ker \phi}:\ker \phi \rightarrow \ker \psi$ is a surjection, again because $\phi = \chi \circ \psi$. The composition of the two maps is $0$ by definition of the kernel of $\chi$. Finally, if $a \in \ker \phi$ and $a$ lies in the kernel of the right map, then by definition $\chi(a) = 0$.
\end{proof}

@@

\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ so that 
\begin{align}
\label{align:pm-span}
\sum_{i=0}^{n} c_i \cdot \bida_i \cdot f_i = 0 && \text{ and } &&c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}

\noindent
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=1}^n f_i^{c_i} $ such that
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor && \text{ for } && i > m
\end{align}

@@

\begin{proof}
By definition of $H^0(\bp^m,dD)$, functions satisfying conditions 
~\eqref{align:pm-span} lie in $H^0(\bp^m,dD)$. To complete the proof, it suffices to check functions satisfying conditions ~\eqref{align:pm-basis} form a basis of $H^0(\bp^m,dD)$. Note that there $\binom{m+ \lfloor dD \rfloor }{m}$ functions satisfying condition ~\eqref{align:pm-basis}. However we know $h^0(\bp^m,dD) = \binom{m+ \lfloor dD \rfloor }{m},$ so it suffices to show that those functions satisfying condition ~\eqref{align:pm-basis} are independent. This follows from the assumption that $f_0,\ldots, f_n$ form a basis of degree 1 rational functions, and so degree $\lfloor d \deg D \rfloor $ monomials in $f_0,\ldots, f_n$ form a basis of degree $\lfloor d \deg D \rfloor $ rational functions.
\end{proof}

\todo{the following bound is the generation bound beginning from evan's paper}

@@

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
$E$ satisfies the conditions of proposition \todo{cite earlier 
proposition}, so by applying that proposition $R_D$ is generated in 
degree at most
\[
	\sum_{i\in T_=(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{i\in T_+(D) \atop
	j \in T_-(D)} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}

@@

\begin{proof}
We first claim that $\psi$ has a relation of the form
\begin{equation}\label{eqn:relations-psi}
	(y_j - \beta_j)\prod_{i=1}^n {y_i}^{c_{i}}
\end{equation}

@@

Note that we can view the relations from Equation
\ref{eqn:relations-psi} as an element of $S$ given
by $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$.  Let $S'\subseteq (\mathbb{R}_{\ge 0})^{n+1}$ be the
set of $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$ satisfying $c_i \ge -\alpha_i d$,
$\bida_j + \sum_{i = 1}^m \bida_i c_i = 0$, and $\bidb_j + \sum_{i=1}^n \bidb_i c_i
= 0$.  A careful computation deduces that $\gamma (d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots, c_n)$ is of the form 
\[
	E + \sum_{k\in T_=} r_k e_k + \sum_{h\in T_+ \atop k\in T_-} r_{h,k} e_{h,k},
\] 
where $E$ is $\mathbb{R}_{\ge 0}$
combination of elements in 
\[
	\{e_i \mbox{ if } i\in S\} \cup \{e_{i,j}: j\in S_- \mbox{ if } i \in S_+\} \cup \{e_{j,i}: j\in S_+ \mbox{ if } i \in S_-\}
\]
given by lattice point $(d_E, w_1, \ldots, w_n)$ with $w_i = 1$.  These conditions imply that 
\[
	d_E \le \left\{ \begin{split}  \ell_i \gcd(\bida_i, \bidb_i) \mbox{ if } i \in S_= 
			\\ \max_{j\in S_-}(\ell_{i,j} (\bida_i \bidb_
j - \bida_j \bidb_i)) \mbox{ if } i\in S_+ \\
			\\ \max_{j\in S_+}(\ell_{j, i} (\bida_j \bidb_
i - \bida_i \bidb_j)) \mbox{ if } i \in S_- \end{split}  \right.
\]

Let $M$ be the set of monomials of $\beta_j$.  Then for $m\in M$, the corresponding lattice point is of the form
\[
	E_m + \sum_{k\in T_=} r_k e_k + \sum_{h\in T_+ \atop k\in T_-} r_{h,k} e_{h,k}
\]
where $E_m$ is a $\mathbb{R}_{\ge 0}$ combination of elements in
\[
	\{e_j : j\in T_= \cap \{1, 2, 3, 4\}\} \cup \{e_{j, k}: j\in T_= \cap \{1, 2, 3, 4\}, k\in T_-\} \cup \{e_{k,j}: k\in T_+, j\in T_- \cap \{1, 2, 3, 4\}\}.
\]
